\subsubsection{\halo}\label{halodata}

A compendium of 1495 transcriptome profiles were collated from a wide
array of experiments conducted by our lab that cover dynamic
transcriptional responses to varied growth, nutritional, and stress
conditions, temperatures, salinities, metal ions, and genetic
perturbations (full set of annotations available online). 1159 of
these are published (Baliga et al., 2004; Baliga et al., 2002; Bonneau
et al., 2007; Facciotti et al., 2010; Facciotti et al., 2007; Kaur et
al., 2006; Kaur et al., 2010; Schmid et al., 2011; Schmid et al.,
2007; Schmid et al., 2009; Whitehead et al., 2006; Whitehead et al.,
2009). 336 are new for this study. Experimental protocols are
identical to (Bonneau et al., 2007). These data, including expression
levels (log$_2$ ratios vs. reference samples) and experimental metadata,
are available as a tab-delimited spreadsheet.

\paragraph{Data normalization} %% note this probably doesnt need to be a separate section

\subsubsection{\eco}\label{ecodata}

The \eco data set was obtained from the DISTILLER website (Lemmens
et al., 2009). These data were collated from publicly available
microarray data consisting of 3 major microarray databases: Stanford
Microarray Database (Demeter et al., 2007), Gene Expression Omnibus
(Barrett et al., 2007) and ArrayExpress (Parkinson et al., 2007), The
experiments cover a range of conditions, including varying carbon
sources, pH, oxygen, metals and temperature.

\paragraph{Data normalization}
